Parametricity in the Presence of GADTs

### 2021-03-04 20-gadts/GADT-story.tex
almost final gadt-story

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2021}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

\begin{abstract}
{\color{red} FIX ME!} There are two ways to represent/interpret
GADTs. They are not the same, and how we view GADTs has profound
consequences for how we program with, and reason about, them.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
{\color{blue} map vs. fmap throughout.} {\color{blue} Add polynomial
  throughout?}
{\color{blue} We say "naturality consequences of parametricity"
when in fact we have naturality without having any parametricity at all.
Change the wording to indicate that naturality follows from
functoriality of the interp.}

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
semantically equivalent, i.e., both have the same interpretation in
the model. Models in which the Church encodings and primitive
representations of all algebraic data types (ADTs) are semantically
equivalent include the operationally-based model of~\cite{pit98,pit00}
and the categorical models of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
Correctness of short cut fusion, and thus semantic equivalence of
Church encodings and primitive representations, was recently extended
to the class of what might be called ``hereditarily polynomial''
{\color{blue} to include truly nested such types} nested types via the
categorical model of~\cite{jgj21}. But, as shown in
Section~\ref{sec:par} below, for some GADTs --- even polynomial ones
--- no such isomorphism holds.

@@

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. To be used in this way, a GADT must be viewed in terms of its
primitive representation. In fact, the functorial completion of a GADT
inherent in its primitive representation is exactly what is needed to
ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
interpreted as a true categorical functor, and can therefore be made
an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
however, that since the functorial completion of a GADT includes
{\color{blue} only?} those data elements that are derivable via
repeated applications of $\mathsf{map}$ starting from those specified
by its syntax, primitive GADTs need not, in general, be nested
types. For example, the functorial completion of the GADT $\mathsf{G}$
defined by
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
\end{array}
\end{equation}
Here, $\mathsf{1}$ is the unit type, whose single element is also
denoted by $\mathsf{1}$, includes elements at any instance
$\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the empty type
$\mathsf{0}$ since there is always a function from $\mathsf{1}$ to
such an $\mathsf{a}$, but does not include any terms at instance
$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
syntactic specification of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$.

@@

Since the shape of an ADT structure --- i.e., a structure whose type
is an instance of an ADT --- is independent of the type of data it
contains, ADTs can be defined polymorphically. As a result, an ADT
structure containing data of type $\mathsf{a}$ can be transformed into
another ADT structure of the exact same shape containing data of
another type $\mathsf{b}$ simply by applying a given function
$\mathsf{f \,:\,a \to b}$ to each of its elements. Indeed, every ADT
$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
class by defining a type-and-data-uniform, structure-preserving,
data-changing $\mathsf{fmap_T}$ function for it.  Moreover, given a
type-independent way of rearranging an ADT structure's shape
$\mathsf{T\,a}$ into the shape for another ADT structure
$\mathsf{T'\,a}$, we get the same structure of type $\mathsf{T'\,b}$
regardless of whether we first rearrange the original structure of
type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$ and then use
$\mathsf{fmap_{T'}}$ to convert that resulting structure to one of
type $\mathsf{T'\,b}$, or we first use $\mathsf{fmap_T}$ to convert
the original structure of type $\mathsf{T\,a}$ to one of type
$\mathsf{T\,b}$ and then rearrange that resulting structure into one
of type $\mathsf{T'\,b}$. For example, if $\mathsf{f : a \to b}$,
$\mathsf{xs \,:\,List\,a}$, and $\mathsf{g : List\,a\to Tree\,a}$
arranges lists into trees in a type-independent way, then we have the
following rearrange-transform property:
\[\mathsf{fmap_{Tree}\,f\,(g \, xs) \,=\, g\,(fmap_{List}\,f\,xs)}\]

@@

Because the types of the ``abstract constructors'' for an ADT are
uniform in their argument types, it is always possible to Church
encode the type constructors themselves as well. For example, the
Church encodings of the type constructors $\mathsf{List}$ and
$\mathsf{Tree}$ are
\[\mathsf{List} = \mathsf{\forall a. \forall b.\,b \to (a \to b \to b)
  \to b}\] 
and 
\[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
  a \to b \to b) \to b}\] respectively. This is what allows an ADT's
associated type constructor to be made an instance of Haskell's
$\mathsf{Functor}$ class. For example, Haskell's built-in
$\mathsf{map}$ function makes the type constructor $\mathsf{List}$ an
instance of the $\mathsf{Functor}$ class, and for $\mathsf{Tree}$ we
can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,x\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
Note, however, that the Church encoding of an ADT carries with it no
expectation whatsoever that such a type-and-data-uniform,
structure-preserving, data-changing $\mathsf{map}$ function can be
defined. It therefore need not satisfy any rearrange-transform
property either.

@@

These kinds of fixpoint equations are entirely sensible at the level
of types. But to ensure that the syntactic fixpoint representing an
ADT actually denotes a semantic object computed as a real-live
fixpoint, the semantic fixpoint calculation must converge.  If, as is
typical, we interpret our types as sets, then this means that the
fixpoint being taken must be of a {\em functor} on the category
$\mathit{Set}$ of sets and functions between them, rather than of a
mere function between sets~\cite{tfca}.  That is, the function $F$
interpreting the type constructor $\mathsf{F}$ constructing the body
of a syntactic fixpoint must not only have an action on sets, but must
also have a {\em functorial action} on functions between
sets. Reflecting this requirement back into syntax gives that
$\mathsf{F}$ must support a suitable\footnote{The $\mathsf{map}$
  function for $\mathsf{F}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}$ --- i.e.,
  preservation of identity functions and composition of functions ---
  even though there is no mechanism in Haskell for enforcing this.}
$\mathsf{map}$ function, i.e., must be an instance of Haskell's
$\mathsf{Functor}$ class.

Requiring $F$ to be a functor will ensure that the interpretation $\mu
F$ of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure
that $\mu F$ is itself a functor, so that the type constructor
$\mathsf{T}$ associated with the ADT $\mathsf{T\,a}$ supports a
$\mathsf{map}$ function as well, we can require that $F$ also be the
fixpoint $\mu H$ of a functor $H$. In this case, however, $H$ must be
a functor on the category $\mathit{Set}^\mathit{Set}$ of functors and
natural transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that
$\mathsf{T\,a} = \mathsf{(\mu H)\,a}$
for the ``type constructor constructor''
$\mathsf{H}$ that supports suitable\footnote{The $\mathsf{map}$
  function for $\mathsf{H}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}^\mathit{Set}$ ---
  i.e., preservation of identity natural transformations and
  composition of natural transformations --- and the $\mathsf{map}$
  function for $\mathsf{H\,F}$ should satisfy syntactic reflections of
  the functoriality properties in $\mathit{Set}$ as in the previous
  footnote. Again, there is no mechanism in Haskell for enforcing
  this.} $\mathsf{map}$ functions.

%class HFunctor f where
%  ffmap :: Functor g => (a -> b) -> f g a -> f g b
%  hfmap :: (g :~> h) -> f g :~> f h
%
%  type (~>) f g = forall x. f x -> g x

@@

In a parametric model, short cut fusion can be used to show that the
Church encoding and the primitive representation of an ADT are
isomorphic as sets. Note, however, that whereas it is impossible even
to {\em state} the rearrange-transform property for Church encodings
of ADTs {\color{blue} unless we first define $\mathsf{map}$ functions
  for them}, this property is simply the reflection back into syntax
of the instance of naturality for the type-independent function that
rearranges structures of type $\mathsf{T\,a}$ into ones of type
$\mathsf{T'\,a}$ and the content-transforming functions
$\mathsf{map_T\,f}$ and $\mathsf{map_{T'}\,f}$ for a function
$\mathsf{f :: a \to b}$, where $\mathsf{T}$ and $\mathsf{T'}$ are the
type constructors associated with these ADTs, respectively.

@@

GADTs relax the restriction on the type instances appearing in a data
type definition. The special form of GADTs known as {\em nested types}
allow the data constructors of a GADT to take as arguments data whose
types involve type instances of the GADT other than the one being
defined. For example, the definition
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the GADT $\mathsf{Seq}$ given in~\eqref{eq:seq}. Note that since the
return type of the data constructor $\mathsf{Pair}$ is not of the form
$\mathsf{Seq\,a}$ for any variable $\mathsf{a}$, $\mathsf{Seq}$ is a
GADT that is not a nested type.

@@

@@

The syntax of GADTs is consistent with programmers' use of them in
situations where different behaviors at different type instances is
desired. This is achieved by allowing the programmer to give the type
signatures of the different data constructors independently, and then
using pattern matching to force the desired type refinement.  Typical
applications of GADTs include generic programming, modeling
programming languages via higher-order abstract syntax, maintaining
invariants in data structures, and expressing constraints in embedded
domain-specific languages. They have also been used, e.g., to
implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

Allowing non-variable type arguments in the return types of a GADT's
data constructors establishes a strong connection between a GADT's
shape and the data it contains. With ADTs, we first choose the shape
of the container and then fill that container with data of whatever
type we like. Critically, the choice of shape is independent of the
data to be stored. With GADTs, however, the shape of the container may
actually {\em depend} on (the type of) the data to be contained. For
example, $\mathsf{Const}$ can create data of any shape
$\mathsf{Seq\,a}$, but $\mathsf{Pair}$ can produce data of shape
$\mathsf{Seq\,a}$ only if $\mathsf{a}$ is a pair type. As a result,
modifying the data in a GADT may actually change the shape of --- or
may even destroy! --- the GADT.

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, it is
essential that a GADT calculus support an equality type
$\mathsf{Equal}$ that is a singleton set when its two type arguments
are the same and is the empty set otherwise. That is, the type
$\mathsf{Equal}$ must be the syntactic reflection of semantic equality
function $\mathit{Equal}$. The type $\mathsf{Equal}$ can either be
defined via GADT syntax, as in
\[\begin{array}{l}
\mathsf{data\, Equal\,a\,b\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}\\
\end{array}\]
\noindent
or, equivalently, represented by its Church encoding
\[\mathsf{Equal\,a\,b} = \mathsf{\forall f.\, (\forall c.\,
  f\,c\,c) \to f\,a\,b}\] in a higher-kinded calculus such as
$F_\omega$~\cite{atk12}. Following the technique in Section~4.5
of~\cite{atk12}, we can represent the GADT $\mathsf{Seq}$ as its
Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\,(\forall d. \,(f\,d + \exists b
  c.\, f\, b \times f \,c \times \mathsf{Equal}\, d \,(b \times c))
  \to f \,d) \to f\, a}\]
This is logically equivalent to the more intuitive encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall d.\,d \to f\, d)
  \to (\forall b\,c\,d.\,Equal\,d\,(b\times c) \to f\,b \to f\,c \to
  f\,d) \to f\,a}\]

@@

Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
varies in the instances of $\mathsf{Seq}$ appearing in the types of
its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
modeled as the fixpoint of any first-order functor. As shown
in~\cite{jp19}, it can, however, be modeled as a solution to the
higher-order fixpoint equation
\[H\,f\,a = a \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
c d. f c \times f d)\,a\] {\color{red} Check this equation against
  next displayed one below.}  where $Lan_K \,F$ is the left Kan
extension of the functor $F$ along the functor $K$. In general, the
left Kan extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal
  C} \to {\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best
functorial approximation to $F$ that factors through $K$. Here, ``best
functorial approximation'' means that $Lan_K \,F$ is the smallest
functor that both extends the image of $K$ to $\cal D$ and agrees with
$F$ on $\cal C$, in the sense that, for any other such functor $H$,
there is a morphism of functors (i.e., a natural transformation) from
$Lan_K \,F$ to $H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $F : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $F$ along $K$} is a
functor $\mathit{Lan}_K\,F : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : F \to (\mathit{Lan}_K\,F) \circ K$ such
that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : F \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,F \to H$ such that
$(\delta K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,F}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

@@

@@

{\color{blue} Is G or G\,a the GADT?}

>>>>

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{Haskell Symposium} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2021}
\acmMonth{8}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

\begin{abstract}
GADTs can be represented either by their Church encodings {\em \`a la}
Atkey or as primitive data types {\em \`a la} Johann and Polonsky.  A
GADT represented as a Church encoding need not have a functorial
semantics, i.e., need not support a $\mathsf{map}$ function. By
contrast, the primitive prepresentation of a GADT must have a
functorial semantics even to be well-defined. The two representations
of a GADT thus need not be the same in general. In this paper we show
that the choice of representation for GADTs has profound consequences
for their parametricity properties, and thus for how we program with,
and reason about, these advanced data types.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}

There are two standard ways to represent data types when studying
modern functional languages: via their Church encodings in a (possibly
higher-kinded) extension of System F~\cite{gir72}, the calculus at the
core of all such languages, and by augmenting System F
%a Hindley-Milner style calculus
with primitives for constructing them directly as fixpoints.
%{\color{blue} Something about when each of these representations is
%  used. The former perhaps more closely represents how data types are
%  represented in implementations, but the latter are how they have
%  heretofore primarily been treated theoretically.}
In models that validate the program optimization known as {\em short
  cut fusion}~\cite{glp93} the two representations are, in fact,
semantically equivalent, i.e., both have the same interpretation in
the model. Models in which the Church encodings and primitive
representations of all algebraic data types (ADTs) are semantically
equivalent include the operationally-based model of~\cite{pit98,pit00}
and the categorical models of~\cite{joh02,joh03}.
%\footnote{{\color{blue} Pitts proves the isomorphism only for
%    polynomial ADTs, whereas Johann {\em et al.}  prove it for any
%    data type that is the fixpoint of a first-order functor.}}
Correctness of short cut fusion, and thus semantic equivalence of
Church encodings and primitive representations, was recently extended
to the class of what might be called ``hereditarily polynomial''
nested types via the categorical model of~\cite{jgj21}. But, as shown
in Section~\ref{sec:par} below, for some GADTs --- even polynomial
ones --- no such isomorphism holds.

@@

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. To be used in this way, a GADT must be viewed in terms of its
primitive representation. In fact, the functorial completion of a GADT
inherent in its primitive representation is exactly what is needed to
ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
interpreted as a true categorical functor, and can therefore be made
an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
however, that since the functorial completion of a GADT includes those
data elements that are derivable via repeated applications of
$\mathsf{map}$ starting from those specified by its syntax, primitive
GADTs need not, in general, be nested types. For example, let
$\mathsf{1}$ be the unit type, whose single element is also denoted by
$\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined by
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
\end{array}
\end{equation}
Then the functorial completion of $\mathsf{G}$ includes elements at
any instance $\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the
empty type $\mathsf{0}$ (there is always a function from $\mathsf{1}$
to such an $\mathsf{a}$), but does not include any terms at instance
$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
syntactic specification of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$.

@@

Since the shape of an ADT structure --- i.e., a structure whose type
is an instance of an ADT --- is independent of the type of data it
contains, ADTs can be defined polymorphically. As a result, an ADT
structure containing data of type $\mathsf{a}$ can be transformed into
another ADT structure of the exact same shape containing data of
another type $\mathsf{b}$ simply by applying a given function
$\mathsf{f \,:\,a \to b}$ to each of its elements. Indeed, every ADT
$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
class by defining a type-and-data-uniform, structure-preserving,
data-changing function $\mathsf{map_T}$ for it.\footnote{We write
  $\mathsf{map_f}$, or simply $\mathsf{map}$ when $\mathsf{f}$ is
  clear from context, for the function $\mathsf{fmap :: (a \to b) \to
    (f\,a \to f\,b)}$ witnessing that a type constructor $\mathsf{f}$
  is an instance of Haskell's $\mathsf{Functor}$ class. This
  emphasizes that $\mathsf{fmap}$ functions in Haskell are intended to
  {\color{blue} satisfy syntactic reflections of the functoriality
    properties --- i.e., preservation of identity functions and
    composition of functions --- even though}
%    there is no mechanism in  Haskell for enforcing this}.}
%
%  be functorial --- i.e., to preserve identity functions and function
%  composition --- even though satisfaction of these properties
this is not
  enforced by the compiler and is instead left to the good intentions
  of the programmer.}
Moreover, given a type-independent way of rearranging an ADT
structure's shape $\mathsf{T\,a}$ into the shape for another ADT
structure $\mathsf{T'\,a}$, we get the same structure of type
$\mathsf{T'\,b}$ regardless of whether we first rearrange the original
structure of type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$
and then use $\mathsf{map_{T'}}$ to convert that resulting structure
to one of type $\mathsf{T'\,b}$, or we first use $\mathsf{map_T}$ to
convert the original structure of type $\mathsf{T\,a}$ to one of type
$\mathsf{T\,b}$ and then rearrange that resulting structure into one
of type $\mathsf{T'\,b}$. For example, if $\mathsf{f : a \to b}$,
$\mathsf{xs \,:\,List\,a}$, and $\mathsf{g : List\,a\to Tree\,a}$
arranges lists into trees in a type-independent way, then we have the
following rearrange-transform property:
\[\mathsf{map_{Tree}\,f\,(g \, xs) \,=\, g\,(map_{List}\,f\,xs)}\]

@@

Because the types of the ``abstract constructors'' for an ADT are
uniform in their argument types, it is always possible to Church
encode the type constructors themselves as well. For example, the
Church encodings of the type constructors $\mathsf{List}$ and
$\mathsf{Tree}$ are
\[\mathsf{List} = \mathsf{\forall a. \forall b.\,b \to (a \to b \to b)
  \to b}\] 
and 
\[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
  a \to b \to b) \to b}\] respectively. This is what allows an ADT's
associated type constructor to be made an instance of Haskell's
$\mathsf{Functor}$ class. For example, Haskell's built-in
$\mathsf{map}$ function makes the type constructor $\mathsf{List}$ an
instance of the $\mathsf{Functor}$ class, and for $\mathsf{Tree}$ we
can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,x\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
Note, however, that the Church encoding of an ADT carries with it no
expectation whatsoever that such a type-and-data-uniform,
structure-preserving, data-changing $\mathsf{map}$ function can be
defined (although one always can). It therefore need not satisfy any
rearrange-transform property either (although it always does if such a
$\mathsf{map}$ function is defined for it).

@@

These kinds of fixpoint equations are entirely sensible at the level
of types. But to ensure that the syntactic fixpoint representing an
ADT actually denotes a semantic object computed as a real-live
fixpoint, the semantic fixpoint calculation must converge.  If, as is
typical, we interpret our types as sets, then this means that the
fixpoint being taken must be of a {\em functor} on the category
$\mathit{Set}$ of sets and functions between them, rather than of a
mere function between sets~\cite{tfca}.  That is, the function $F$
interpreting the type constructor $\mathsf{F}$ constructing the body
of a syntactic fixpoint must not only have an action on sets, but must
also have a {\em functorial action} on functions between
sets. Reflecting this requirement back into syntax gives that
$\mathsf{F}$ must support a suitable
%\footnote{The $\mathsf{map}$
%  function for $\mathsf{F}$ should satisfy syntactic reflections of
%  the functoriality properties in $\mathit{Set}$ --- i.e.,
%  preservation of identity functions and composition of functions ---
%  even though there is no mechanism in Haskell for enforcing this.}
$\mathsf{map}$ function, i.e., must be an instance of Haskell's
$\mathsf{Functor}$ class.

Requiring $F$ to be a functor ensures that the interpretation $\mu F$
of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure that
$\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
associated with the ADT $\mathsf{T\,a}$ also supports a $\mathsf{map}$
function, we can require that $F$ also be the fixpoint $\mu H$ of a
functor $H$. In this case, however, $H$ must be a functor on the
category $\mathit{Set}^\mathit{Set}$ of functors and natural
transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for
the ``type constructor constructor'' $\mathsf{H}$ that supports
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ should
  satisfy syntactic reflections of the functoriality properties in
  $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
  natural transformations and composition of natural transformations
  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ should
  satisfy syntactic reflections of the functoriality properties in
  $\mathit{Set}$ as in the previous footnote. Again, there is no
  mechanism in Haskell for enforcing this.} $\mathsf{map}$ functions.

@@

\vspace*{0.05in}

In models that validate short cut fusion for ADTs and nested types,
the Church encoding and the primitive representation of an ADT or
nested type are isomorphic as sets. But whereas it is impossible even
to {\em state} the rearrange-transform property for {\color{blue} raw
  --- i.e., $\mathsf{map}$less ---} Church encodings of such data
types, such a property is simply the reflection back into syntax of
the instance of naturality for the type-independent function that
rearranges structures of type $\mathsf{T\,a}$ into ones of type
$\mathsf{T'\,a}$ and the content-transforming functions
$\mathsf{map_T\,f}$ and $\mathsf{map_{T'}\,f}$ for a function
$\mathsf{f :: a \to b}$, where $\mathsf{T}$ and $\mathsf{T'}$ are the
type constructors associated with these ADTs, respectively.

@@

GADTs relax the restriction on the type instances appearing in a data
type definition. The special form of GADTs known as {\em nested types}
allow the data constructors of a GADT to take as arguments data whose
types involve type instances of the GADT other than the one being
defined. For example, the definition
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructors $\mathsf{PLeaf :: a \to PTree\,a}$
and $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the GADT $\mathsf{Seq}$ given in~\eqref{eq:seq}. Note that since the
return type of the data constructor $\mathsf{Pair}$ is not of the form
$\mathsf{Seq\,a}$ for any variable $\mathsf{a}$, $\mathsf{Seq}$ is a
GADT that is not a nested type.

@@

GADTs are used in precisely those situations in which different
behaviors at different instances of a data type are desired. This is
achieved by allowing the programmer to give the type signatures of the
GADT's data constructors independently, and then using pattern
matching to force the desired type refinement. Applications of GADTs
include generic programming, modeling programming languages via
higher-order abstract syntax, maintaining invariants in data
structures, and expressing constraints in embedded domain-specific
languages. GADTs have also been used, e.g., to implement tagless
interpreters~\cite{pl04,pr06,pvww06}, to improve memory
performance~\cite{min15}, and to design APIs~\cite{pen20}.

@@

The syntax of GADTs allows non-variable type arguments in the return
types of their data constructors. This establishes a strong connection
between a GADT's shape and the data it contains. With ADTs, we first
choose the shape of the container and then fill that container with
data of whatever type we like; critically, the choice of shape is
independent of the data to be stored. With GADTs, however, the shape
of the container may actually {\em depend} on (the type of) the data
to be contained. For example, $\mathsf{Const}$ can create data of any
shape $\mathsf{Seq\,a}$, but $\mathsf{Pair}$ can produce data of shape
$\mathsf{Seq\,a}$ only if $\mathsf{a}$ is a pair type. As a result,
modifying the data in a GADT's container may actually change the shape
of --- or may even destroy! --- that container.

To determine the possible shapes of a GADT's container we must
pattern-match on the type of the data to be contained. For this, it is
essential that a GADT calculus support an equality type
$\mathsf{Equal}$ that is a singleton set when its two type arguments
are the same and is the empty set otherwise. That is, the type
$\mathsf{Equal}$ must be the syntactic reflection of semantic equality
function $\mathit{Equal}$. The type $\mathsf{Equal}$ can either be
defined via GADT syntax, as in
\[\begin{array}{l}
\mathsf{data\, Equal\,a\,b\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,c\,c}\\
\end{array}\]
\noindent
or, equivalently, represented by its Church encoding
\[\mathsf{Equal\,a\,b} = \mathsf{\forall f.\, (\forall c.\,
  f\,c\,c) \to f\,a\,b}\] in a higher-kinded calculus such as
$F_\omega$~\cite{atk12}. Following the technique in Section~4.5
of~\cite{atk12}, we can represent the GADT $\mathsf{Seq}$ as its
Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\,(\forall d. \,(f\,d + \exists b
  c.\, f\, b \times f \,c \times \mathsf{Equal}\, d \,(b \times c))
  \to f \,d) \to f\, a}\]
This is logically equivalent to the more intuitive encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall d.\,d \to f\, d)
  \to (\forall b\,c\,d.\,Equal\,d\,(b\times c) \to f\,b \to f\,c \to
  f\,d) \to f\,a}\]

@@

Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
varies in the instances of $\mathsf{Seq}$ appearing in the types of
its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
modeled as the fixpoint of any first-order functor. As shown
in~\cite{jp19}, it can, however, be modeled as a solution to the
higher-order fixpoint equation
\[H\,f\,b = b \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
c d. f c \times f d)\,b\] where $Lan_K \,F$ is the left Kan extension
of the functor $F$ along the functor $K$. In general, the left Kan
extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal C} \to
{\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best functorial
approximation to $F$ that factors through $K$. Here, ``best functorial
approximation'' means that $Lan_K \,F$ is the smallest functor that
both extends the image of $K$ to $\cal D$ and agrees with $F$ on $\cal
C$, in the sense that, for any other such functor $H$, there is a
morphism of functors (i.e., a natural transformation) from $Lan_K \,F$
to $H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $F : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $F$ along $K$} is a
functor $\mathit{Lan}_K\,F : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : F \to (\mathit{Lan}_K\,F) \circ K$ such
that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : F \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,F \to H$ such that
$(\delta K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,F}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

@@

\vspace*{0.05in}

@@
